---
title: Exploring classes to define Experiments
author: Eunice Jun
output: html_document
---

General notes: 
- could change the OOP-esque nature of API

# "Target" program 
```{r}
exp = Experiment()
exp.total_num_trials = 4 # total number of trials per participant; use this to expand alternate/sequence/random trials?

# TASKS
c1 = task("c1")
c2 = task("c2")
e1 = task("e1")
e2 = task("e2")

# Group based on type of task
creation_tasks = TaskGroup(c1, c2)
editing_tasks = TaskGroup(e1, e2)
exp.add_task_group(creation_tasks) # minor: should be "set_task.../task_group..."?
exp.add_task_group(editing_tasks) # minor: should be "set_task.../task_group..."?

# Group based on task number
group_1 = TaskGroup(c1, e1)
group_2 = TaskGroup(c2, e2) 

# CONDITIONS
ffl = condition("ffl")
latex = condition("latex")
conditions = conditions(ffl, latex) 
exp.add_conditions(conditions) # minor: should be "set_conditions"?
# Note: There could be hierarchies of conditions, to form ConditionGroup (i.e., Barr et al. linguistics)

# TRIALS: Constraints on what trials are considered valid 

# All creation tasks should be completed before all editing tasks
s = sequence(creation_tasks, editing_tasks) # should these return an expanded list for all trials?
exp.enforce_order(s)
# Trials should alternate between ffl and latex
a = alternate(ffl, latex)
exp.enforce_order(a)
# TODO: The below isn't quite right:
# Group 1 tasks and Group 2 tasks should alternate -- 
# but it's really not about alternation as much as it is about enforcing same order in creation and editing
a = alternate(group1, group2)

visualize(exp) # output table

# OTHER PROGRAMS TO WRITE: 
# - With vs. Without ordering based on task group!
```

# Functions to provide
visualize every building block 

# Class design notes

CONSIDER: Group could be a "generic" concept that applies to Tasks and Conditions
- QUESTION: Should these task groups exist in the Experiment class or in a separate Task Hierarchy class???
- What if the Task Hierarchy is an instance variable of the Experiment class?


# Experiment Class (main class)
```{r}
# Define Experiment class
setClass(
  Class = "Experiment",
  slots = list(
    tasks = "list",
    conditions = "list",
    trials = "list" # for now -- likely want to have own class/data structure
  )
)

# Define constructor
Experiment <- function() {
  
  # TODO: Set default values for the pivs
  # Validate and create an instance of the class
  new("Experiment", tasks=list(), conditions=list(), trials=list())
}

# toString for Experiment
# @prints tasks in this Experiment
setMethod(
  f = "show",
  signature = "Experiment",
  definition = function(object) {
    cat("Experiment Object:\n")
    cat("tasks:", unlist(object@tasks), "\n")
    cat("conditions:", unlist(object@conditions), "\n")
    cat("trials:", unlist(object@trials), "\n")
  }
)

visualize = function() {
    # construct table

    # read in file and show PDF in-line (ideal)
}


```

# Task Class
```{r}
# Define Task class
setClass(
  Class = "Task",
  slots = list(
    name = "char",
  )
)

# Define constructor
Task <- function(name) {
  # Validate and create an instance of the class
  new("Task", name = name)
}

# @returns Task object
# @param name of task
task = function(name) {
    return(Task(name))
}

# toString for Task
# @prints tasks in this Task
setMethod(
  f = "show",
  signature = "Task",
  definition = function(object) {
    cat("Task Object:\n")
    cat("name:", unlist(object@name), "\n")
  }
)
```

# Task Group Class
```{r}
# Define TaskGroup class
setClass(
  Class = "TaskGroup",
  slots = list(
    name = "list", # specify that this is a list of Task objects?
  )
)
# Define constructor
TaskGroup <- function(tasks) {
  # Validate and create an instance of the class
  new("TaskGroup", tasks = tasks)
}

# @return TaskGroup
# @param number of tasks in this group can vary
# TODO: overload function so that parameters can be Tasks or str/chars
# TODO: Should this be the TaskGroup constructor?
task_group = function(...) {
    tasks = list()
    for(t in ...) {
        tasks = append(tasks, t)
    }

    return(TaskGroup(tasks))
}

# toString for TaskGroup
# @prints tasks in this TaskGroup
setMethod(
  f = "show",
  signature = "TaskGroup",
  definition = function(object) {
    cat("TaskGroup Object:\n")
    cat("tasks:", unlist(object@tasks), "\n")
  }
)

# TODO: Does this need to be setMethod???
visualize <- function(taskGroup) {
    # Show table
}
```

# Condition Class
```{r}
# Specify conditions
# Define Condition class
setClass(
  Class = "Condition",
  slots = list(
    name = "char", # Is there anything else that conditions contain besides a name?
  )
)
# Define constructor
Condition <- function(name) {
  # Validate and create an instance of the class
  new("Condition", name = name)
}

# @returns Condition object
# @param name of task
condition = function(name) {
    return(Condition(name))
}
```

# Trial Classes
```{r}
# Define Trial class
setClass(
  Class = "SingleTrial",
  slots = list(
    task = Task,
    condition = Condition,
  )
)
# Define constructor
SingleTrial <- function(task, condition) {
  # Validate and create an instance of the class
  new("SingleTrial", task = task, condition = condition)
}

# Define OrderedTrials class: Describes all the possible orderings of trials(?)
setClass(
  Class = "OrderedTrials",
  slots = list(
    trials = list, # list of SingleTrial
    rules = list # list of Rules (should this be a class?) ????
  )
)

# NOT SURE THIS IS WHAT I WANT:
# Define constructor
OrderedTrials <- function(trials) {
  # Validate and create an instance of the class
  new("OrderedTrials", trials)
}

# Part of OrderedTrials class?
rules = function() {

}
```

# Trial specification/construction
Core of trial specification is constraints => Work through what the constraints look like

Check out [trials-focused doc](constraints.md)

===
# What If

summarize(exp.trials.rules): summarizes rules in NL? So that we can more easily check what we're doing is what we intend / generate text for paper?
exp.trials.rules # output list of rules for generating trials encoded so far

---
a = alternate()
s = sequence()
r = random()

exp.trials.enforce_order(a, s) 

Trials class
Trial
===